---
title: "A&E Waiting Times"
output:
  html_document:
    toc: yes
    toc_float: true
    toc_depth: '2'
    highlight: tango
    df_print: paged
---

A&E Data Mart User Guide
https://www.isdscotland.org/Health-Topics/Emergency-Care/Emergency-Department-Activity/Data-Quality/AE2-User-Guide_V1-3.pdf

There are two types of data submitted to the A&E datamart: episode and aggregate level
data. Episode level data is now submitted from all Emergency Departments in Scotland,
and a growing number of community A&E departments and Minor Injury Units. Small A&E
sites may submit aggregate data which contains only the number of attendances and the
number who spent 4 hours or less in the department. This data can only be used to
monitor monthly activity and the compliance with the 4 hour standard. 

Two types of file can be submitted to the A&E data mart;
Episode – a detailed file containing a record for each patient
Aggregate – a summary file containing a record for each location

**Important information to include in data bias and reliability section of report**
https://publichealthscotland.scot/publications/ae-activity-and-waiting-times/ae-activity-and-waiting-times-month-ending-30-november-2022/
How these different urgent and emergency care services operate and the way space such as trolleyed areas are used can vary both by hospital (depending on the facilities available and how services have been set up to meet local demand) and also over time (depending on how many people need urgent or emergency care at different times). These differences could impact on the consistency and comparability of A&E statistics and it is important to bear this in mind when comparing statistics for different areas or hospitals, or looking at trend data over time.

**Data Dictionary**
https://www.opendata.nhs.scot/dataset/monthly-emergency-department-activity-and-waiting-times/resource/2a4adc0a-e8e3-4605-9ade-61e13a85b3b9

Attendances are only reported at aggregate level for minor injury units and smaller A&E departments. 
The columsn labelled QF in the data give the symbol which explains why data is missing (I think). The qualifier "z" is entered for most (if not all) the missing data in this dataset and it means "not applicable".

number_meeting_target_aggregate - the number of attendances that are admitted, transferred or discharged wthin 4 hours of arrival at aggregate level

discharge_destination_admission_to_same - The number of attendances that were admitted to the same NHS heathcare provider/hospital. (Episode level data only)

discharge_destination_other_speciality - The number of attendances that were discharged to a Private healthcare provider and the number of attendances that died. (Episode level data only)

discharge_destination_residence - 	The number of attendances that were discharged to a Private residence, residential institution or temporary residence. (Episode level data only)

discharge_destination_transfer - The number of attendances that were transferred to another NHS healthcare provider/hospital. (Episode level data only)

discharge_destination_unknown - The number of attendances where the discharge destination was Other or Not Known. (Episode level data only)

## Libraries and Setting up
```{r}
library(tidyverse)
library(lubridate)
```

```{r}
source(here::here("palette_theme/define_theme.R"))
source(here::here("palette_theme/plot_timeseries.R"))

```

```{r}
waiting_times <- read_csv(here::here("clean_data/a_and_e_data_clean.csv")) %>% 
  janitor::clean_names()
```


When you filter by "Minor Injury Unit or Other" the percentages are close to 100% across all years. Filtering by "Emergency Department" shows a terrible decline in the % of patients being seen within 4 hours. 




Calculate 2018 / 2019 average percent meeting target
```{r}
# Adding percent meeting target column to dataframe
waiting_times <- waiting_times %>% 
  mutate(percent_meeting_target = number_meeting_target_aggregate / 
              number_of_attendances_aggregate * 100)  

# Calculate average percent meeting target across all rows for 2018 and 2019
avg_2018_2019 <- waiting_times %>%  
  filter(date >= "2018-01-01" & date <= "2019-12-31") %>% 
  summarise(average_percent_meeting_target = mean(percent_meeting_target))
```

Average percent meeting target across 2018 and 2019 is 96.66%

```{r}
waiting_times %>% 
#  filter(department_type == "Emergency Department") %>% 
#  filter(hbt == "Highlands") %>% 
  filter(date >= "2020-01-01") %>% 
  group_by(date) %>% 
  summarise(percent_meeting_target_by_month = mean(percent_meeting_target)) %>% 
  
  ggplot(aes(date, percent_meeting_target_by_month)) +
  geom_line(fill = "grey60") +
  geom_hline(yintercept = 95, colour = "red", linetype = "dashed") + 
  geom_hline(yintercept = 97, colour = "blue", linetype = "dashed") +
# use coord_cartesian to zoom in on y axis not scale_y_continuous() or ylim()
  coord_cartesian(ylim=c(75,100)) +
  theme_phs +
  labs(title = "Percentage of Attendances Meeting 4 Hour Target",
       subtitle = "Data Averaged by Month",
       y = "% Meeting 4 Hour Target",
       x = "Date") +
  geom_rect(aes(xmin = ymd("2022-9-15"), xmax = ymd("2023-03-15"), 
                ymin = 0, ymax = Inf), fill = "skyblue", alpha = 0.01) +
  geom_rect(aes(xmin = ymd("2021-9-15"), xmax = ymd("2022-03-15"), 
                ymin = 0, ymax = Inf), fill = "skyblue", alpha = 0.01) +
  geom_rect(aes(xmin = ymd("2020-9-15"), xmax = ymd("2021-03-15"), 
                ymin = 0, ymax = Inf), fill = "skyblue", alpha = 0.01) +
  geom_rect(aes(xmin = ymd("2020-01-01"), xmax = ymd("2020-03-15"), 
                ymin = 0, ymax = Inf), fill = "skyblue", alpha = 0.01) 
#  geom_rect(aes(xmin = ymd("2018-9-15"), xmax = ymd("2019-03-15"), 
#                ymin = 0, ymax = Inf), fill = "skyblue", alpha = 0.01) +
#  geom_rect(aes(xmin = ymd("2017-12-15"), xmax = ymd("2018-03-15"), 
#                ymin = 0, ymax = Inf), fill = "skyblue", alpha = 0.01)


```

```{r}
monthly_target_data <- waiting_times %>% 
#  filter(department_type == "Emergency Department") %>% 
#  filter(hbt == "Highlands") %>% 
  filter(date >= "2020-01-01") %>% 
  group_by(date) %>% 
  summarise(percent_meeting_target_by_month = mean(percent_meeting_target))



timeseriesplot(monthly_target_data, aes(date, percent_meeting_target_by_month),
               "Percentage of Attendances Meeting 4 Hour Target", 
               "% Meeting 4 Hour Target") +
  labs(subtitle = "Data Averaged By Month") +
  theme(axis.text = element_text(size = 12)) +
  # Make sure only whole percent number shows on y-axis (not e.g. 92.5%)
  scale_y_continuous(labels = scales::label_number(accuracy = 1)) +
  geom_hline(yintercept = 95, colour = "red", linetype = "dashed") + 
  geom_hline(yintercept = 97, colour = "blue", linetype = "dashed") +
  annotate(geom = "text", x = as.Date("2022-08-01"), y = 94.5, 
           label = "NHS 95% Target", colour = "red") +
    annotate(geom = "text", x = as.Date("2022-08-01"), y = 96.5, 
           label = "2018/2019 Average", colour = "blue")
  
```


```{r}
timeseriesplot <- function(plotdata,plotmapping,plottitle,plotylabel){
  plotlim <- as.Date(c("2020-01-01","2022-12-31")) 
  ggplot(plotdata,plotmapping) +
    geom_line() +    
    # add dashed lines to show the boundaries 
    geom_vline(xintercept=ymd(20201001),color="gray",linetype="dotted", linewidth = 0.5) +                                   geom_vline(xintercept=ymd(20211001),color="gray",linetype="dotted", linewidth = 0.5) + 
    geom_vline(xintercept=ymd(20221001),color="gray",linetype="dotted", linewidth = 0.5) + 
    geom_vline(xintercept=ymd(20200401),color="gray",linetype="dotted", linewidth = 0.5) + 
    geom_vline(xintercept=ymd(20210401),color="gray",linetype="dotted", linewidth = 0.5) + 
    geom_vline(xintercept=ymd(20220401),color="gray",linetype="dotted", linewidth = 0.5) +
    scale_x_date(limits=plotlim, date_breaks="3 month", labels = scales::label_date_short(), expand = c(0,0)) +
#    scale_y_continuous(expand = c(0,0)) +
    xlab("Date") +   
    ylab(plotylabel) +
    ggtitle(plottitle) +  
    theme_phs +
    #add rectangles to denote winter
    annotate("rect", 
             xmin=c(ymd(plotlim[1]),ymd(20201001),ymd(20211001),ymd(20221001)),
             xmax=c(ymd(20200401),ymd(20210401),ymd(20220401),plotlim[2]), 
             ymin=c(-Inf,-Inf,-Inf,-Inf), 
             ymax=c(Inf,Inf,Inf,Inf), 
             alpha=0.1, fill="gray")
}

```










