---
title: "A&E Discharge Data"
output:
  html_document:
    toc: yes
    toc_float: true
    toc_depth: '2'
    highlight: tango
    df_print: paged
---


**Data Dictionary**
https://www.opendata.nhs.scot/dataset/monthly-emergency-department-activity-and-waiting-times/resource/2a4adc0a-e8e3-4605-9ade-61e13a85b3b9

Attendances are only reported at aggregate level for minor injury units and smaller A&E departments. 
The columsn labelled QF in the data give the symbol which explains why data is missing (I think). The qualifier "z" is entered for most (if not all) the missing data in this dataset and it means "not applicable".

number_meeting_target_aggregate - the number of attendances that are admitted, transferred or discharged wthin 4 hours of arrival at aggregate level

discharge_destination_admission_to_same - The number of attendances that were admitted to the same NHS heathcare provider/hospital. (Episode level data only)

discharge_destination_other_speciality - The number of attendances that were discharged to a Private healthcare provider and the number of attendances that died. (Episode level data only)

discharge_destination_residence - 	The number of attendances that were discharged to a Private residence, residential institution or temporary residence. (Episode level data only)

discharge_destination_transfer - The number of attendances that were transferred to another NHS healthcare provider/hospital. (Episode level data only)

discharge_destination_unknown - The number of attendances where the discharge destination was Other or Not Known. (Episode level data only)

## Libraries and Setting up
```{r}
library(tidyverse)
library(lubridate)
```

```{r}
theme_fc <- function(){
  
  theme_minimal() %+replace%   
    
    theme(
      axis.line = element_line(colour = "grey50"),
      axis.ticks = element_line(colour = "grey50"),
      plot.title = element_text(size = 16, 
                                face = "bold", 
                                colour = "grey30",
                                margin = margin(b = 12),
                                hjust = 0),
      plot.subtitle = element_text(size = 14,  
                                   colour = "grey30",
                                   margin = margin(b = 12),
                                   hjust = 0),
      axis.title = element_text(size = 12, 
                                face = "bold", 
                                colour = "grey30"),
      axis.title.x = element_text(margin = margin(t = 12)),
      # chaning the margin on y-axis causes it to flip to horizontal   
      # axis.title.y = element_text(margin = margin(b = 12)),
      axis.text = element_text(size = 11, 
                               colour = "grey30"),
      legend.title = element_blank(),
      legend.text = element_text(size = 11, 
                                 colour = "grey30")
    )
}

plot_colours <- c("#446e9b", "#999999", "#3cb521", "#d47500", "#cd0200", "#3399f3",
                           "#333333", "#6610f2", "yellow", "brown", "peachpuff", "#6f42c1",
                           "#e83e8c", "#fd7e14", "#20c997", "#000000", "grey50", "#eeeeee")
```



```{r}
data <- read_csv("/Users/fionacarson/Documents/codeclan_work/codeclan_work/week_08/codeclan_group_project/data/Monthly A&E activity and waiting times.csv")

data <- janitor::clean_names(data)
```

## Cleaning and Wrangling Data

```{r}
data <- data %>% 
  rename(date = month) %>% 
#  mutate(year = str_extract(date, "[0-9]{4}"),
#         month = str_sub(date, -2, -1)) %>% 
  mutate(date = ym(date)) %>% 
    mutate(year = year(date),
         month = month(date)) 
  
data <- data %>%   
  select(-ends_with("qf")) %>% 
  filter(date >= "2018-01-01")

```



```{r}
discharge_data <- data %>% 
  select(-id, -country, -starts_with("num"), -starts_with("attend"))
```


## Discharge Data

### Discharge Destination
```{r}
discharge_data %>% 
  pivot_longer(discharge_destination_admission_to_same:discharge_destination_unknown,
    names_to = "discharge_destination", values_to = "number_of_people") %>% 
  group_by(date, discharge_destination) %>% 
  summarise(total = sum(number_of_people, na.rm = TRUE)) %>% 
  
  ggplot(aes(date, total)) +
  geom_line(aes(colour = discharge_destination)) +
  scale_y_continuous(limits = c(0, 120000), n.breaks = 10)
  



```
Was wondering if this data might show an increase in people being admitted to hospital in recent times but this is not showing in the data. I suppose if we look at percentage numbers it might show something but it doesn't seem to be an obvious trend so best to leave for now. 


Representing data on a stacked bar chart

```{r}
discharge_data %>% 
  pivot_longer(discharge_destination_admission_to_same:discharge_destination_unknown,
    names_to = "discharge_destination", values_to = "number_of_people") %>% 
  group_by(year, discharge_destination) %>% 
  summarise(total = sum(number_of_people, na.rm = TRUE)) %>% 
  
  ggplot(aes(year, total)) + 
  geom_col(aes(fill = discharge_destination), position = "fill")
  




```

I don't think there is anything interesting here to explore further