---
title: "A&E Discharge Data"
output:
  html_document:
    toc: yes
    toc_float: true
    toc_depth: '2'
    highlight: tango
    df_print: paged
---


**Data Dictionary**
https://www.opendata.nhs.scot/dataset/monthly-emergency-department-activity-and-waiting-times/resource/2a4adc0a-e8e3-4605-9ade-61e13a85b3b9

Attendances are only reported at aggregate level for minor injury units and smaller A&E departments. 
The columsn labelled QF in the data give the symbol which explains why data is missing (I think). The qualifier "z" is entered for most (if not all) the missing data in this dataset and it means "not applicable".

number_meeting_target_aggregate - the number of attendances that are admitted, transferred or discharged wthin 4 hours of arrival at aggregate level

discharge_destination_admission_to_same - The number of attendances that were admitted to the same NHS heathcare provider/hospital. (Episode level data only)

discharge_destination_other_speciality - The number of attendances that were discharged to a Private healthcare provider and the number of attendances that died. (Episode level data only)

discharge_destination_residence - 	The number of attendances that were discharged to a Private residence, residential institution or temporary residence. (Episode level data only)

discharge_destination_transfer - The number of attendances that were transferred to another NHS healthcare provider/hospital. (Episode level data only)

discharge_destination_unknown - The number of attendances where the discharge destination was Other or Not Known. (Episode level data only)

## Libraries and Setting up
```{r}
library(tidyverse)
library(lubridate)
```

```{r}
theme_fc <- function(){
  
  theme_minimal() %+replace%   
    
    theme(
      axis.line = element_line(colour = "grey50"),
      axis.ticks = element_line(colour = "grey50"),
      plot.title = element_text(size = 16, 
                                face = "bold", 
                                colour = "grey30",
                                margin = margin(b = 12),
                                hjust = 0),
      plot.subtitle = element_text(size = 14,  
                                   colour = "grey30",
                                   margin = margin(b = 12),
                                   hjust = 0),
      axis.title = element_text(size = 12, 
                                face = "bold", 
                                colour = "grey30"),
      axis.title.x = element_text(margin = margin(t = 12)),
      # chaning the margin on y-axis causes it to flip to horizontal   
      # axis.title.y = element_text(margin = margin(b = 12)),
      axis.text = element_text(size = 11, 
                               colour = "grey30"),
      legend.title = element_blank(),
      legend.text = element_text(size = 11, 
                                 colour = "grey30")
    )
}

plot_colours <- c("#446e9b", "#999999", "#3cb521", "#d47500", "#cd0200", "#3399f3",
                           "#333333", "#6610f2", "yellow", "brown", "peachpuff", "#6f42c1",
                           "#e83e8c", "#fd7e14", "#20c997", "#000000", "grey50", "#eeeeee")
```



```{r}
data <- read_csv("/Users/fionacarson/Documents/codeclan_work/codeclan_work/week_08/codeclan_group_project/data/Monthly A&E activity and waiting times.csv")

data <- janitor::clean_names(data)
```

## Cleaning and Wrangling Data

```{r}
data <- data %>% 
  rename(date = month) %>% 
#  mutate(year = str_extract(date, "[0-9]{4}"),
#         month = str_sub(date, -2, -1)) %>% 
  mutate(date = ym(date)) %>% 
    mutate(year = year(date),
         month = month(date)) 
  
data <- data %>%   
  select(-ends_with("qf")) %>% 
  filter(date >= "2018-01-01")

```



```{r}
discharge_data <- data %>% 
  select(-id, -country, -starts_with("num"), -starts_with("attend"))
```


## Discharge Data

### Discharge Destination
```{r}
discharge_data %>% 
  pivot_longer(discharge_destination_admission_to_same:discharge_destination_unknown,
    names_to = "discharge_destination", values_to = "number_of_people") %>% 
  group_by(date, discharge_destination) %>% 
  summarise(total = sum(number_of_people, na.rm = TRUE)) %>% 
  
  ggplot(aes(date, total)) +
  geom_line(aes(colour = discharge_destination)) +
  scale_y_continuous(limits = c(0, 120000), n.breaks = 10)
  



```


```{r}

filter(date > "2020-01-01") %>% 
  group_by(date) %>% 
  summarise(attendances_agg = sum(number_of_attendances_aggregate, na.rm = TRUE),
            attendances_episode = sum(number_of_attendances_episode, na.rm = TRUE)) %>% 
  ggplot() +
  geom_line(aes(date, attendances_agg)) +
  geom_line(aes(date, attendances_episode), colour = "blue") +
  labs(title = "Number of Attendances Aggregate (black) and Episodes (blue)",
       y = "Number of Attendances") +
  scale_y_continuous(limits = c(0, 160000), n.breaks = 10)
```


Use aggregate numbers in first place as there is more data for this
Only discharge numbers for episodes though 

### 2018/2019 averages by month
```{r}
avg_2018_2019 <- waiting_times %>% 
  filter(date < "2020-01-01") %>% 
  group_by(date) %>%
  summarise(total = sum(number_meeting_target_aggregate)) %>%
# Tried tidying this up by creating year and month columns in original waiting_times dataframe but they disappear during the summarise function above. Am sure there must be a way to dry this code out. 
  mutate(year = year(date),
         month = month(date)) %>% 
  group_by(month) %>% 
  summarise(avg_2018_2019 = mean(total))
```


Adding 2018/2019 averages into waiting times dataframe 
```{r}
waiting_times %>% 
  filter(date >= "2020-01-01") %>% 
  group_by(date) %>% 
  summarise(num_meeting_target_by_month = sum(number_meeting_target_aggregate)) %>% 
  mutate(year = year(date),
         month = month(date), .after = date) %>% 
  left_join(avg_2018_2019, by = "month") %>% 
  mutate(percent_variation = ((num_meeting_target_by_month - avg_2018_2019)/avg_2018_2019)*100) %>% 
  
  ggplot(aes(date, percent_variation)) + 
  geom_col() +
  theme_fc() + 
  labs(title = "Percentage Variation in Number Meeting 4 Hour 
       Target (Attendance Number)",
       subtitle = "Compared to 2018 / 2019 Average")


```

Q Is it valid to say the number meeting target was lower if the volume of people overall was much lower? Should it all be done in percentage?
A No because it is the number that needs dealt with and A&E should be able to meet targets BUT the graph above isn't telling us anything useful.


### Percent meeting target

```{r}
waiting_times <- waiting_times %>% 
  mutate(percent_meeting_target = number_meeting_target_aggregate / 
           number_of_attendances_aggregate * 100)

waiting_times %>% 
#  filter(department_type == "Emergency Department") %>% 
  group_by(date) %>% 
  summarise(percent_meeting_target_by_month = mean(percent_meeting_target)) %>% 
  ggplot(aes(date, percent_meeting_target_by_month)) +
  geom_col(fill = "grey60") +
  geom_hline(yintercept = 95, colour = "red", linetype = "dashed") + 
  theme_fc() +
  labs(title = "Percentage of Attendances Meeting 4 Hour Target",
       subtitle = "Data Averaged by Month",
       y = "% Meeting 4 Hour Target",
       x = "Date") +
  geom_rect(aes(xmin = ymd("2020-03-01"), xmax = ymd("2020-06-01"), 
                ymin = 0, ymax = Inf), 
                fill = "orange", alpha = 0.01)

```
When you filter by "Minor Injury Unit or Other" the perecentages are close to 100% across all years. Filtering by "Emergency Department" shows a terrible decline in the % of patients being seen within 4 hours. 


### Percentages compared to 2018 / 2019 averages




```{r}
waiting_times %>% 
  filter(date >= "2020-01-01") %>% 
  group_by(date) %>% 
  summarise(num_meeting_target_by_month = sum(number_meeting_target_aggregate)) %>% 
  mutate(year = year(date),
         month = month(date), .after = date) %>% 
  left_join(avg_2018_2019, by = "month") %>% 
  mutate(percent_variation = ((num_meeting_target_by_month - avg_2018_2019)/avg_2018_2019)*100) %>% 
  
  ggplot(aes(date, percent_variation)) + 
  geom_col() +
  theme_fc() + 
  labs(title = "Percentage Variation in Number Meeting 4 Hour 
       Target (Attendance Number)",
       subtitle = "Compared to 2018 / 2019 Average")

```












```{r}
discharge_data <- data %>% 
  select(date, hbt, treatment_location, department_type, 
         number_of_attendances_aggregate, number_of_attendances_episode,
         number_meeting_target_aggregate, number_meeting_target_episode,
         discharge_destination_admission_to_same,
         discharge_destination_other_specialty,
         discharge_destination_residence,
         discharge_destination_transfer,
         discharge_destination_unknown) %>% 
    mutate(total_discharge = sum(discharge_destination_admission_to_same,
         discharge_destination_other_specialty,
         discharge_destination_residence,
         discharge_destination_transfer,
         discharge_destination_unknown, na.rm = TRUE))
```









```{r}
discharge_data %>% 
  filter(date > "2019-01-01") %>% 
  group_by(date) %>% 
  summarise(admitted_same_hospital = 
              discharge_destination_admission_to_same / total_discharge, 
            residence = 
              discharge_destination_residence / total_discharge, 
            transfer = 
              discharge_destination_transfer / total_discharge, 
            other_speciality_inc_death = 
              discharge_destination_other_specialty / total_discharge,
            unknown = 
              discharge_destination_unknown / total_discharge) %>% 
  ggplot() +
  geom_line(aes(date, admitted_same_hospital)) +
  geom_line(aes(date, residence), colour = "blue") +
  labs(title = "Number Meeting 4 h Target Aggregate (black) and Episodes (blue)",
       y = "Number Meeting Target") 
 # scale_y_continuous(limits = c(0, 160000), n.breaks = 10)
```

