---
title: "Waiting Times"
output:
  html_document:
    toc: yes
    toc_float: true
    toc_depth: '2'
    highlight: tango
    df_print: paged
---


Stage of treatment waiting times
https://www.opendata.nhs.scot/dataset/stage-of-treatment-waiting-times
See link for data dictionary

There is data for completed and ongoing waits. The issue with using completed waits is that it is skewed towards faster treatment for more urgent cases. I suppose the opposited could be said for ongoing waits. So which to use. Possible to use both?

Median - Median wait (days) of patients seen within the quarter
90th_percentile - 90th percentile wait (days) of patients seen within the quarter

The data includes all the health boards plus all of scotland (no data recorded for this for some quarters in 2017 and 2018). The Jubilee hospital also has a separate code - not clear if its data is also included under Glasgow health board. 


```{r}
library(tidyverse)
```

# Completed Waits

```{r}
completed_waits <- read_csv("/Users/fionacarson/Documents/codeclan_work/codeclan_work/week_08/codeclan_group_project/downloaded_data/sot_performance_completed_waits_sep22.csv")

completed_waits <- janitor::clean_names(completed_waits)
```
## Cleaning Data

```{r}
completed_waits <- completed_waits %>% 
#  rename(date = month) %>% 
#  mutate(year = str_extract(date, "[0-9]{4}"),
#         month = str_sub(date, -2, -1)) %>% 
  mutate(quarter_ending = ymd(quarter_ending)) %>% 
    mutate(year = year(quarter_ending),
         month = month(quarter_ending)) 

completed_waits <- completed_waits %>%   
  select(-ends_with("qf"))
#  filter(quarter_ending >= "2018-01-01")
```

## Number of Patients Seen

```{r}
completed_waits %>% 
  filter(!hbt == "S92000003") %>% 
  group_by(quarter_ending) %>% 
  summarise(total_seen_by_date = sum(number_seen, na.rm = TRUE)) %>% 
  ggplot(aes(quarter_ending, total_seen_by_date)) + 
  geom_line()

```
What is going on with this data?? Why the huge dip from Q2 2017 to Q1 2019??

## Median number of days waited

```{r}
completed_waits %>% 
  group_by(quarter_ending) %>% 
  summarise(mean_median = mean(median, na.rm = TRUE)) %>% 
  ggplot(aes(quarter_ending, mean_median)) + 
  geom_line() +
  labs(title = "Median Number of Days Waited",
       x = "Date",
       y = "Median Number of Days Waited")
```

## Number waiting over 12 weeks

```{r}
completed_waits %>% 
  group_by(quarter_ending) %>% 
  summarise(total_over_12w = sum(waited_over12weeks, na.rm = TRUE)) %>% 
  ggplot(aes(quarter_ending, total_over_12w)) + 
  geom_line() +
  labs(title = "Number of Patients Waiting Over 12 Weeks",
       x = "Date",
       y = "Number of Patients")
```

# Ongoing Waits

```{r}
ongoing_waits <- read_csv("/Users/fionacarson/Documents/codeclan_work/codeclan_work/week_08/codeclan_group_project/downloaded_data/sot_performance_ongoing_waits_sep22.csv")

ongoing_waits <- janitor::clean_names(ongoing_waits)
```
```{r}
ongoing_waits <- ongoing_waits %>% 
#  rename(date = month) %>% 
#  mutate(year = str_extract(date, "[0-9]{4}"),
#         month = str_sub(date, -2, -1)) %>% 
  mutate(month_ending = ymd(month_ending)) %>% 
    mutate(year = year(month_ending),
         month = month(month_ending)) 

ongoing_waits <- ongoing_waits %>%   
  select(-ends_with("qf"))
#  filter(quarter_ending >= "2018-01-01")
```



## Number of Patients Seen

```{r}
ongoing_waits %>% 
  filter(hbt == "SB0801") %>% 
  group_by(month_ending) %>% 
  summarise(total_waiting_by_date = sum(number_waiting, na.rm = TRUE)) %>% 
  ggplot(aes(month_ending, total_waiting_by_date)) + 
  geom_line()
```
```{r}
ongoing_waits %>% 
  distinct(hbt)
```

