---
title: "Waiting Times"
output:
  html_document:
    toc: yes
    toc_float: true
    toc_depth: '2'
    highlight: tango
    df_print: paged
---


Stage of treatment waiting times
https://www.opendata.nhs.scot/dataset/stage-of-treatment-waiting-times
See link for data dictionary

There is data for completed and ongoing waits. The issue with using completed waits is that it is skewed towards faster treatment for more urgent cases. I suppose the opposite could be said for ongoing waits. So which to use. Possible to use both?

Median - Median wait (days) of patients seen within the quarter
90th_percentile - 90th percentile wait (days) of patients seen within the quarter

The data includes all the health boards plus all of scotland (no data recorded for this for some quarters in 2017 and 2018). The Jubilee hospital also has a separate code - not clear if its data is also included under Glasgow health board. 

From report
https://publichealthscotland.scot/media/16971/2022-11-29-wt-ipdcop-report_revised.pdf
Patient Rights (Scotland) Act 2011 which established a 12 week Treatment Time Guarantee (TTG) written into
legislation for eligible patients who are due to receive planned inpatient or day case
treatment from 1 October 2012. The Act states that eligible patients must start to
receive that treatment within 12 weeks (84 days) of the treatment being agreed. This
guarantee is based on completed waits. 

Inpatient, day case, and new outpatient waiting times are also vital components in
the delivery of the 18 Weeks Referral to Treatment standard (a maximum whole
journey waiting time of 18 weeks from referral to treatment).

New outpatients - patients added to the waiting list for their first appointment at a consultant led clinic. PHS do not collect information nationally on waiting times for return outpatients.  

New outpatients national standard - 95% of new outpatients wait no longer than
12 weeks from referral to being seen.

It should be noted that due to the exclusion of eight key diagnostic tests from new
outpatient activity as of October 2019, along with possible shifts in patient pathways
and service provision towards increased use of non-consultant led clinics, a full
return to pre-pandemic levels is unlikely, particularly for the number of patients seen
and additions to the list.

The total Scotland data was created by summing the health boards. This gave the same results as the original Scotland data except it filled in the zero values and increased the low values for 2017 and 2018. This means that the Jubilee data is separate and not included in NHS Glasgow. It is a special separate health board. 

```{r}
library(tidyverse)
library(lubridate)
```

The data for scotland (S92000003) and tayside (S9200030) record zero patients for Q2 2017 to Q2 2018 and low values for Q3-4 2018. 

The dip seen in the summed data must be due to the tip seen in the Tayside data. 

Ongoing waits may be a more representative dataset. It could be looked at as a worst case scenario as these are the less high priority people so they will be waiting longer. It is a snapshot of the current waiting lists.

```{r}
ongoing_waits <- read_csv(here::here("clean_data/treatment_waiting_times_ongoing.csv"))
```

```{r}
palette = read_csv(here::here("palette_theme/phs_palette.csv"))
# load theme
source(here::here("palette_theme/define_theme.R"))
#load plotting function
source(here::here("palette_theme/plot_timeseries.R"))
```


## Number of Patients On Waiting list

The ongoing wait data suffers the same issues as the completed waits. The Scotland and Tayside data are low or zero for a period across 2017 and 2018. 

Graph for reference
```{r}
ongoing_waits %>% 
  filter(!hb == "S92000003") %>% 
  filter(specialty == "Z9") %>% 
  group_by(month_ending) %>% 
  summarise(total = sum(number_waiting, na.rm = TRUE),
            total_12w = sum(number_waiting_over12weeks, na.rm = TRUE)) %>% 
  ggplot(aes(month_ending, total)) + 
  geom_line() +
  geom_line(aes(month_ending, total_12w), colour = "red") +
  scale_y_continuous(labels = scales::comma)


```




Calculating 2018 and 2019 average, ensuring it can be filtered and therefore calculated at health board level.
```{r}
avg_2018_2019 <- ongoing_waits %>% 
  filter(month_ending >= "2018-01-01" & month_ending <= "2019-12-31") %>% 
  filter(!hb == "S92000003") %>% 
# Have to filter by Z9 which is all specialties otherwise each patient counted twice
    filter(specialty == "Z9") %>% 
#  filter(hb == "S08000030") %>% 
  group_by(month_ending) %>% 
  summarise(num_waiting_2018_2019_by_month = sum(number_waiting, na.rm = TRUE),
            num_waiting_over12weeks_2018_2019_by_month = 
              sum(number_waiting_over12weeks, na.rm = TRUE)) %>% 
  summarise(avg_num_waiting = mean(num_waiting_2018_2019_by_month),
            avg_num_waiting_over12weeks = mean(num_waiting_over12weeks_2018_2019_by_month))
```


```{r}
ongoing_waits %>% 
  filter(!hb == "S92000003") %>% 
# Have to filter by Z9 which is all specialties otherwise each patient counted twice
    filter(specialty == "Z9") %>% 
#  filter(hb == "S08000025") %>% 
  group_by(month_ending) %>% 
  mutate(total_waiting_by_month = sum(number_waiting, na.rm = TRUE),
         total_waiting_over12weeks_by_month = sum(number_waiting_over12weeks, 
                                                    na.rm = TRUE)) %>% 
 mutate(percentage_var = (total_waiting_by_month - avg_2018_2019$avg_num_waiting)
            / avg_2018_2019$avg_num_waiting * 100,
        percentage_var_12w = (total_waiting_over12weeks_by_month - 
                                avg_2018_2019$avg_num_waiting_over12weeks) /
          avg_2018_2019$avg_num_waiting_over12weeks * 100) %>% 
  
  timeseriesplot(aes(month_ending, percentage_var), "Treatment Waiting Times", 
                 "% Variation Compared to 2018/2019") 
#   geom_line(aes(month_ending, percentage_var_12w), colour = "red")



```

Chose this metric to display as it is simple and gives a strong simple message. The number of patients waiting for treatment is increasing. Shows a bottleneck.


