---
title: "Waiting Times"
output:
  html_document:
    toc: yes
    toc_float: true
    toc_depth: '2'
    highlight: tango
    df_print: paged
---


Stage of treatment waiting times
https://www.opendata.nhs.scot/dataset/stage-of-treatment-waiting-times
See link for data dictionary

There is data for completed and ongoing waits. The issue with using completed waits is that it is skewed towards faster treatment for more urgent cases. I suppose the opposite could be said for ongoing waits. So which to use. Possible to use both?

Median - Median wait (days) of patients seen within the quarter
90th_percentile - 90th percentile wait (days) of patients seen within the quarter

The data includes all the health boards plus all of scotland (no data recorded for this for some quarters in 2017 and 2018). The Jubilee hospital also has a separate code - not clear if its data is also included under Glasgow health board. 


```{r}
library(tidyverse)
library(lubridate)
```

# Completed Waits

```{r}
completed_waits <- read_csv("/Users/fionacarson/Documents/codeclan_work/codeclan_work/week_08/codeclan_group_project/downloaded_data/sot_performance_completed_waits_sep22.csv")

completed_waits <- janitor::clean_names(completed_waits)
```
## Cleaning Data

```{r}
completed_waits <- completed_waits %>% 
#  rename(date = month) %>% 
#  mutate(year = str_extract(date, "[0-9]{4}"),
#         month = str_sub(date, -2, -1)) %>% 
  mutate(quarter_ending = ymd(quarter_ending)) %>% 
    mutate(year = year(quarter_ending),
         month = month(quarter_ending)) 

completed_waits <- completed_waits %>%   
  select(-ends_with("qf"))
#  filter(quarter_ending >= "2018-01-01")
```

## Number of Patients Seen

```{r}
completed_waits %>% 
  group_by(quarter_ending, hbt) %>% 
  summarise(total_seen_by_date = sum(number_seen, na.rm = TRUE)) %>% 
  ggplot(aes(quarter_ending, total_seen_by_date)) + 
  geom_line(aes(colour = hbt))
```
The data for scotland (S92000003) and tayside (S9200030) record zero patients for Q2 2017 to Q2 2018 and low values for Q3-4 2018. It is not clear why at this point. 

If the values from each health board are summed does this match the scotland values?

```{r}
totals_by_summing_hb <- completed_waits %>% 
  filter(!hbt == "S92000003") %>% 
  group_by(quarter_ending) %>% 
  summarise(totals = sum(number_seen, na.rm = TRUE))

scot_data_only <- completed_waits %>% 
  filter(hbt == "S92000003") %>% 
  group_by(quarter_ending) %>% 
  summarise(totals = sum(number_seen, na.rm = TRUE))

ggplot(scot_data_only, aes(quarter_ending, totals)) +
  geom_line() +
  geom_line(data = totals_by_summing_hb, colour = "red")

```

The dip seen in the summed data must be due to the tip seen in the Tayside data. 

I don't see how we can use this data to create a 2018 - 2018 average. 

## Median number of days waited

The median isn't affected as its just missing data which isn't used to calculate median. 

```{r}
completed_waits %>% 
  group_by(quarter_ending) %>% 
  summarise(mean_median = mean(median, na.rm = TRUE)) %>% 
  ggplot(aes(quarter_ending, mean_median)) + 
  geom_line() +
  labs(title = "Median Number of Days Waited",
       x = "Date",
       y = "Median Number of Days Waited")
```

## Number waiting over 12 weeks

The number waiting over 12 weeks is affected by the dip

```{r}
completed_waits %>% 
  group_by(quarter_ending) %>% 
  summarise(total_over_12w = sum(waited_over12weeks, na.rm = TRUE)) %>% 
  ggplot(aes(quarter_ending, total_over_12w)) + 
  geom_line() +
  labs(title = "Number of Patients Waiting Over 12 Weeks",
       x = "Date",
       y = "Number of Patients")
```




# Ongoing Waits

Ongoing waits may be a more representative dataset. It could be looked at as a worst case scenario as these are the less high priority people. It is a snapshot of the current waiting lists.

```{r}
ongoing_waits <- read_csv("/Users/fionacarson/Documents/codeclan_work/codeclan_work/week_08/codeclan_group_project/downloaded_data/sot_performance_ongoing_waits_sep22.csv")

ongoing_waits <- janitor::clean_names(ongoing_waits)
```
```{r}
ongoing_waits <- ongoing_waits %>% 
#  rename(date = month) %>% 
#  mutate(year = str_extract(date, "[0-9]{4}"),
#         month = str_sub(date, -2, -1)) %>% 
  mutate(month_ending = ymd(month_ending)) %>% 
    mutate(year = year(month_ending),
         month = month(month_ending)) 

ongoing_waits <- ongoing_waits %>%   
  select(-ends_with("qf"))
#  filter(quarter_ending >= "2018-01-01")
```




## Number of Patients On Waiting list

The ongoing wait data suffers the same issues as the completed waits. The scotland and Tayside data are low or zero for a period across 2017 and 2018. 

```{r}
ongoing_waits %>% 
  filter(!hbt == "S92000003") %>% 
#  filter(hbt == "S08000025") %>% 
  group_by(month_ending) %>% 
  summarise(total_waiting_by_date = sum(number_waiting, na.rm = TRUE)) %>% 
  ggplot(aes(month_ending, total_waiting_by_date)) + 
  geom_line()
```

Chose this metric to display as it is simple and gives a strong simple message. The number of patients waiting for treatment is increasing. Shows a bottleneck.


Numbers waiting > 12 weeks

```{r}
ongoing_waits %>% 
  filter(!hbt == "S92000003") %>% 
#  filter(hbt == "S08000025") %>% 
  group_by(month_ending) %>% 
  summarise(total_waiting_over12weeks_by_date = sum(number_waiting_over12weeks, na.rm = TRUE)) %>% 
  ggplot(aes(month_ending, total_waiting_over12weeks_by_date)) + 
  geom_line()

```

This could be a good metric to chose as it shows a relatively static period over 2018 - 2019. 


Percentage waiting > 12 weeks

```{r}
ongoing_waits %>% 
  mutate(percent_over12weeks = number_waiting_over12weeks / number_waiting * 100) %>% 
  group_by(month_ending) %>% 
  summarise(pc_waiting_over12weeks_by_date = mean(percent_over12weeks, na.rm = TRUE)) %>% 
  ggplot(aes(month_ending, pc_waiting_over12weeks_by_date)) + 
  geom_line()

```

