---
title: "Investigating Data - Covid Cases"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

Looking at weekly covid cases
weekly variability split by age bands
note that although file contains health boards, data is not broken down by health board!


```{r}
library("tidyverse")
library("janitor")
library("lubridate")
```

```{r}
hb_codes <- read_csv("health_board_codes.csv")
hb_codes <- clean_names(hb_codes)
```

load in datafiles

```{r}
covid_cases <- read_csv("admissions_ageband_week.csv")
covid_cases <- clean_names(covid_cases)

```

data cleaning

```{r}

```

merge hbnames into datafiles
```{r}
covid_cases <- left_join(covid_cases,hb_codes)
```

do the analysis needed for identifying winter and 'crisis'

load in datafiles

```{r}
covid_cases <- read_csv("admissions_ageband_week.csv")
covid_cases <- clean_names(covid_cases)

```

take a look at what is in dataset

```{r}
covid_cases %>% 
  distinct(hb_name)
```
14 health boards but data not broken down by health board!

this summary code not needed so ignore it.
```{r}
all_covid_cases <- covid_cases %>% 
  group_by(wdate, hb) %>% 
  mutate(all_ages = sum(admissions, na.rm = TRUE)) %>% 
  ungroup() 
  #filter(age_band == "18-29")
```
```{r}
all_covid_cases %>% 
  ggplot()+
  aes(x=wdate, y=all_ages)+
  geom_line()
```

back to dataset - the summary data, for all ages is called Total


can see that covid cases don't have a simple winter/summer pattern - much more complex particularly coming into 2021.

```{r}
covid_cases %>% 
  filter(age_band != "Total") %>% 
ggplot() +
aes(x=wdate, y = admissions, colour = age_band) +
geom_line() 
```
quick look at demographics

```{r}
covid_cases %>% 
  group_by(age_band) %>% 
 summarise(total = sum(admissions, na.rm = TRUE)) %>% 
 arrange(desc(total))
```

boxplot for winter/summer
```{r}
covid_cases %>%
 filter(age_band != "Total") %>%   
 group_by(year, iswinter) %>% 
 ggplot()+
 aes(x=iswinter, y=admissions, fill = hb_name) +  
 geom_col() +
facet_grid(~year)
```


Now look by health boards 


```{r}
covid_cases %>% 
  filter(age_band == "Total") %>% 
  ggplot()+
  aes(x=wdate, y=admissions, colour = hb_name)+
  geom_line()+
  geom_line(aes(x=all_covid_cases$wdate, y =all_covid_cases$all_ages, colour = "red"))
```

