---
title: "Investigating Data - Covid Wider Impacts"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

Looking at wider covid impacts data
weekly variability - related to 2018/19 baseline

revised version - now extracting all Scotland more easily.

```{r}
library("tidyverse")
library("janitor")
library("lubridate")
```

```{r}
hb_codes <- read_csv("health_board_codes.csv")
hb_codes <- clean_names(hb_codes)
```

load in datafiles

```{r}
weekly_admissions_spec <- read_csv("Covid admissions by health board and speciality.csv")
weekly_admissions_spec <- clean_names(weekly_admissions_spec)

weekly_admissions_dep <- read_csv("Covid admissions by health board and deprivation.csv")
weekly_admissions_dep <- clean_names(weekly_admissions_dep)

weekly_admissions_demog <- read_csv("Covid admissions by health board, age and sex.csv")
weekly_admissions_demog <- clean_names(weekly_admissions_demog)
```

data cleaning
note probably more cleaning needed before we finalise
I changed the spelling but it turns out both spellings are correct!

```{r}
weekly_admissions_spec <- weekly_admissions_spec %>% 
  rename("speciality"= "specialty") %>% 
  rename("speciality_qf"= "specialty_qf") 
```

merge hbnames into datafiles
```{r}
weekly_admissions_spec <- left_join(weekly_admissions_spec,hb_codes)
weekly_admissions_demog <- left_join(weekly_admissions_demog,hb_codes)
weekly_admissions_dep <- left_join(weekly_admissions_dep,hb_codes) 
```

do the analysis needed for identifying qinter and 'crisis'

```{r}
weekly_admissions_spec <- weekly_admissions_spec %>% 
  mutate(year = as.integer(str_sub(week_ending,1,4))) %>% 
  mutate(month = as.integer(str_sub(week_ending,5,6))) %>% 
  mutate(day = as.integer(str_sub(week_ending,7,8))) %>% 
  mutate(wdate = ymd(week_ending)) %>% 
  # identify "All Scotland" data
  mutate(hb_name = ifelse(hb=="S92000003","All Scotland",hb_name)) %>% 
  mutate(hb_name = ifelse(is.na(hb_name),"NHS Region Unknown",hb_name)) %>% 
  mutate(iswinter = ifelse(month %in% c(4,5,6,7,8,9),FALSE,TRUE)) %>% 
  mutate(above_thresh = ifelse(percent_variation>0,7,0))
```

```{r}
weekly_admissions_demog <- weekly_admissions_demog %>% 
  mutate(year = as.integer(str_sub(week_ending,1,4))) %>% 
  mutate(month = as.integer(str_sub(week_ending,5,6))) %>% 
  mutate(day = as.integer(str_sub(week_ending,7,8))) %>% 
  mutate(wdate = ymd(week_ending)) %>% 
  # identify "All Scotland" data
  mutate(hb_name = ifelse(hb=="S92000003","All Scotland",hb_name)) %>% 
  mutate(hb_name = ifelse(is.na(hb_name),"NHS Region Unknown",hb_name)) %>% 
  mutate(iswinter = ifelse(month %in% c(4,5,6,7,8,9),FALSE,TRUE)) %>% 
  mutate(above_thresh = ifelse(percent_variation>0,7,0))
```

```{r}
weekly_admissions_dep <- weekly_admissions_dep %>% 
  mutate(year = as.integer(str_sub(week_ending,1,4))) %>% 
  mutate(month = as.integer(str_sub(week_ending,5,6))) %>% 
  mutate(day = as.integer(str_sub(week_ending,7,8))) %>% 
  mutate(wdate = ymd(week_ending)) %>% 
  # identify "All Scotland" data
  mutate(hb_name = ifelse(hb=="S92000003","All Scotland",hb_name)) %>% 
  mutate(hb_name = ifelse(is.na(hb_name),"NHS Region Unknown",hb_name)) %>% 
  mutate(iswinter = ifelse(month %in% c(4,5,6,7,8,9),FALSE,TRUE)) %>% 
  mutate(above_thresh = ifelse(percent_variation>0,7,0))
```

take a look at what is in each dataset

```{r}
weekly_admissions_spec %>% 
  distinct(hb_name)
```
14 health boards plus 'all scotland' and an NA - need to work out what to do with this

```{r}
weekly_admissions_spec %>% 
  distinct(admission_type)
```
Data neatly divided into emergency and planned
```{r}
weekly_admissions_spec %>% 
  distinct(speciality)
```
note that some groupings are also combined. need to take a closer look at what these mean.

```{r}
weekly_admissions_demog %>% 
  distinct(age_group)
```
7 age groups and "all ages"

```{r}
weekly_admissions_demog %>% 
  distinct(sex)
```
just male - female and all

```{r}
weekly_admissions_dep %>% 
  distinct(simd_quintile)
```
note there is no 'all' category in this dataset

plot total admissions - you should get the same overall trends from each dataset
```{r}
weekly_admissions_demog %>% 
  filter(age_group == "All ages") %>% 
  filter(sex =="All") %>% 
  filter(admission_type == "All") %>% 
  filter(hb_name =="All Scotland") %>% 
ggplot() +
aes(x=wdate, y = number_admissions) +
geom_line(colour='red') 
```


```{r}
weekly_admissions_spec %>% 
  filter(speciality == "All") %>% 
  filter(admission_type == "All") %>% 
  filter(hb_name =="All Scotland") %>% 
ggplot() +
aes(x=wdate, y = number_admissions) +
geom_line(colour='red') 
```
same graph as above!

weekly hospital admissions hover around 14000 per week
expect 168,000 per quarter?

pre=pandemic hospital admissions hovered around 15500 per week
expect 186,000 per quarter

```{r}
weekly_admissions_spec %>% 
  filter(speciality == "All") %>% 
  filter(admission_type == "All") %>% 
  filter(hb_name =="All Scotland") %>% 
ggplot() +
aes(x=wdate, y = percent_variation) +
geom_line(colour='red') 
```

Overall(all specialities, all admissions, all healthboards) admissions are only at 90% of pre-pandemic levels


quick look at demographics

```{r}
weekly_admissions_demog %>% 
  filter(sex == "All") %>% 
  filter(age_group != "All ages") %>% 
  filter(hb_name =="All Scotland") %>% 
  filter(admission_type == "All") %>% 
ggplot() +
aes(x=wdate, y = number_admissions, color = age_group) +
geom_line() #+
#scale_colour_brewer(palette = "Set2") 
```
age plot shows some variation between age groups.

quick look at sex differences
```{r}
weekly_admissions_demog %>% 
  filter(sex != "All") %>% 
  filter(age_group == "All ages") %>% 
  filter(hb_name =="All Scotland") %>% 
  filter(admission_type == "All") %>% 
ggplot() +
aes(x=wdate, y = percent_variation, color = sex) +
geom_line() #+
#scale_colour_brewer(palette = "Set2") 
```

quick look at index of deprivation. 1 is most deprived and 5 is least deprived

```{r}
weekly_admissions_dep %>% 
  filter(admission_type == "All") %>% 
  filter(hb_name =="All Scotland") %>% 
ggplot() +
aes(x=wdate, y = percent_variation, color = as.factor(simd_quintile)) +
geom_line() 
```
something happening in 2022 - why is there divergence between 1 and 5?



look at admissions type
```{r}
weekly_admissions_spec %>% 
  filter(admission_type != "All") %>% 
  filter(speciality=="All") %>% 
  filter(hb_name =="All Scotland") %>% 
ggplot() +
aes(x=wdate, y = percent_variation, colour = admission_type) +
geom_line()
```
need to check what the 'spike' in planned at end of 2021 is related to

quick look at specialities
```{r}
weekly_admissions_spec %>% 
  filter(admission_type == "All") %>% 
  filter(speciality !="All") %>% 
  filter(hb_name =="All Scotland") %>% 
ggplot() +
aes(x=wdate, y = percent_variation, colour=speciality) +
geom_line()
```
take a closer look at paediatrics
```{r}
weekly_admissions_spec %>% 
  filter(admission_type == "All") %>% 
  filter(speciality !="All") %>% 
  filter(str_detect(speciality,"Paed")) %>% 
  filter(hb_name =="All Scotland") %>% 
ggplot() +
aes(x=wdate, y = number_admissions, colour=speciality) +
geom_line()
```


```{r}
weekly_admissions_spec %>% 
  filter(admission_type == "All") %>% 
  filter(speciality !="All") %>% 
  filter(str_detect(speciality,"Paed")) %>% 
  filter(hb_name =="All Scotland") %>% 
ggplot() +
aes(x=wdate, y = percent_variation, colour=speciality) +
geom_line()
```
take a closer look at medical

```{r}
weekly_admissions_spec %>% 
  filter(admission_type == "All") %>% 
  filter(speciality !="All") %>% 
  filter(str_detect(speciality,"Medical")) %>% 
  filter(hb_name =="All Scotland") %>% 
ggplot() +
aes(x=wdate, y = percent_variation, colour=speciality) +
geom_line()
```
simpler plot of specialities
```{r}
weekly_admissions_spec %>% 
  filter(admission_type == "Planned") %>% 
  filter(speciality !="Medical (incl. Cardiology & Cancer)") %>% 
  filter(!str_detect(speciality,"Paed")) %>% 
  filter(speciality !="All") %>% 
  filter(speciality !="Accident & Emergency") %>% 
  filter(hb_name =="All Scotland") %>% 
ggplot() +
aes(x=wdate, y = percent_variation, colour=speciality) +
geom_line() +
scale_colour_brewer(palette = "Set1") 
```

```{r}
weekly_admissions_spec %>% 
  filter(admission_type == "Emergency") %>% 
  filter(speciality !="Medical (incl. Cardiology & Cancer)") %>% 
  filter(!str_detect(speciality,"Paed")) %>% 
  filter(speciality !="All") %>% 
  filter(speciality !="Accident & Emergency") %>% 
  filter(hb_name =="All Scotland") %>% 
ggplot() +
aes(x=wdate, y = percent_variation, colour=speciality) +
geom_line() +
scale_colour_brewer(palette = "Set1") 
```
boxplot for different specialities? some values greater than 8000% (i have filtered them out ) indicates we need some more work on data cleaning!

```{r}
weekly_admissions_spec %>% 
  #filter(admission_type == "All") %>% 
  #filter(speciality!="All") %>% 
  #filter(hb_name =="NHS Region Unknown") %>% 
  #group_by(speciality) %>% 
  #need to take a good look at what these values are!
  filter(percent_variation>500) 
```


```{r}
weekly_admissions_spec %>% 
  filter(admission_type == "All") %>% 
  filter(speciality!="All") %>% 
  group_by(speciality) %>% 
  #need to take a good look at what these values are!
  filter(percent_variation<8000) %>% 
  ggplot()+
  aes(x=speciality, y=percent_variation)+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
some specialities have faced higher admission levels than others.


Now look by health boards 

```{r}
all_admissions_byhb <- weekly_admissions %>% 
  filter(admission_type == "All") %>% 
  filter(speciality=="All") %>% 
  group_by(hb_name) %>% 
  summarise(mean_percnt_var = mean(percent_variation),
            min_percnt_var = min(percent_variation),
            max_percnt_var = max(percent_variation)
           ) %>%
  #mutate(all_percent_variation = all_admissions/all_avg20182019)  
  arrange(desc(max_percnt_var))
all_admissions_byhb
```
```{r}
weekly_admissions_spec %>% 
  filter(admission_type == "All") %>% 
  filter(speciality=="All") %>% 
  group_by(hb_name) %>% 
  ggplot()+
  aes(x=wdate, y=percent_variation, colour = hb_name)+
  geom_line()
```

```{r}
weekly_admissions_spec %>% 
  filter(admission_type == "All") %>% 
  filter(speciality=="All") %>% 
  group_by(hb_name) %>% 
  ggplot()+
  aes(x=hb_name, y=percent_variation)+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


for some health boards a significant number of admissions levels are above the levels seen in 2018-19. could we parhaps try and use this as an indicator?

Need to think more carefully about what the data in 'region unknown' means 


```{r}
weekly_admissions_spec %>% 
  filter(admission_type == "All") %>% 
  filter(speciality=="All") %>% 
  ggplot() +
  aes(x=wdate, y=percent_variation, colour=hb_name) +
  geom_line()+
  facet_wrap(~hb_name)
```
How is NA calculated - still need to check??

```{r}
weekly_admissions_spec %>% 
  filter(admission_type == "All") %>% 
  filter(speciality == "All") %>% 
  filter(hb_name != "All Scotland") %>% 
ggplot() +
aes(x=wdate, y = number_admissions, colour = hb_name) +
geom_line()
```
admission values shows all health boards - health boards vary massively in size and admission numbers but mostly same overall pattern.

```{r}
weekly_admissions_spec %>% 
  filter(admission_type == "All") %>% 
  filter(speciality=="Accident & Emergency") %>% 
  #take a good look at what these values are!
  filter(percent_variation<8000) %>% 
  group_by(hb_name) %>% 
  ggplot()+
  aes(x=hb_name, y=percent_variation)+
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
do only certain health boards have an A+E dept?

```{r}
weekly_admissions_spec %>% 
  filter(admission_type == "Emergency") %>% 
  #filter(speciality=="Accident & Emergency") %>% 
  #take a good look at what these values are!
  filter(percent_variation<8000) %>% 
  group_by(hb_name) %>% 
  ggplot()+
  aes(x=hb_name, y=percent_variation)+
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

first attempt at a 'crisis' calculation

```{r}
weekly_admissions %>% 
  filter(admission_type == "All") %>% 
  filter(speciality=="All") %>% 
  #filter(admission_type =="Emergency") %>% 
  #filter(speciality=="Accident & Emergency") %>% 
  #filter(year == 2020) %>% 
  filter(iswinter) %>%
  group_by(hb_name) %>% 
  summarise(count_bad_days = sum(above_thresh), 
            count_days = n()*7, 
            pcnt_bad_days = 100*(sum(above_thresh)/(n()*7)), 
            mean_percentvar =mean(percent_variation)) %>%
  ggplot() +
  aes(x=hb_name, y=mean_percentvar) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
weekly_admissions_spec %>% 
  filter(speciality=="All") %>% 
  filter(admission_type =="All") %>% 
  #filter(year == 2021) %>% 
  #filter(iswinter) %>%
  group_by(hb_name) %>% 
  summarise(count_bad_days = sum(above_thresh), 
            count_days = n()*7, 
            pcnt_bad_days = 100*(sum(above_thresh)/(n()*7)), 
            mean_percentvar =mean(percent_variation)) %>%
  ggplot() +
  aes(x=hb_name, y=pcnt_bad_days) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
weekly_admissions_spec %>% 
  filter(speciality=="All") %>% 
  filter(admission_type =="Emergency") %>% 
  #filter(speciality=="Accident & Emergency") %>% 
  #filter(year == 2020) %>% 
  filter(iswinter) %>%
  group_by(hb_name) %>% 
  summarise(count_bad_days = sum(above_thresh), 
            count_days = n()*7, 
            pcnt_bad_days = 100*(sum(above_thresh)/(n()*7)), 
            mean_percentvar =mean(percent_variation)) %>%
  ggplot() +
  aes(x=hb_name, y=pcnt_bad_days) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
take a look by health board

```{r}
weekly_admissions_spec %>% 
  filter(speciality=="All") %>% 
  filter(admission_type =="Emergency") %>% 
  #filter(speciality=="Accident & Emergency") %>% 
  #filter(year == 2020) %>% 
  #filter(iswinter) %>%
  group_by(hb_name) %>% 
  summarise(count_bad_days = sum(above_thresh), 
            count_days = n()*7, 
            pcnt_bad_days = 100*(sum(above_thresh)/(n()*7)), 
            mean_percentvar =mean(percent_variation)) %>%
  ggplot() +
  aes(x=hb_name, y=pcnt_bad_days) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


quick look at winter v summer

```{r}
weekly_admissions_spec %>% 
  filter(speciality=="Accident & Emergency") %>% 
  filter(admission_type =="All") %>% 
  group_by(iswinter, year) %>% 
  summarise(pcnt_bad_days = sum(above_thresh)/n()*7, mean_percentvar =mean(percent_variation)) 
```
  no obvious big discrepancy in winter admissions for this parameter
  winters of 2020 and 2022 are 'worse' than summers, percentages are small and maybe not          significant? 

```{r}
weekly_admissions_spec %>% 
  filter(speciality=="All") %>% 
  filter(admission_type =="All") %>% 
  group_by(iswinter, year) %>% 
  summarise(count_bad_days = sum(above_thresh), 
            count_days = n()*7, 
            pcnt_bad_days = 100*(sum(above_thresh)/(n()*7)), 
            mean_percentvar =mean(percent_variation)) %>% 
  ggplot() +
  aes(x=iswinter, y=pcnt_bad_days) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(~year)
```
```{r}
weekly_admissions_spec %>% 
  filter(speciality=="All") %>% 
  filter(admission_type =="Emergency") %>% 
  group_by(iswinter, year) %>% 
  summarise(count_bad_days = sum(above_thresh), 
            count_days = n()*7, 
            pcnt_bad_days = 100*(sum(above_thresh)/(n()*7)), 
            mean_percentvar =mean(percent_variation)) %>% 
  ggplot() +
  aes(x=iswinter, y=pcnt_bad_days) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(~year)
```
```{r}
weekly_admissions_spec %>% 
  filter(speciality=="Accident & Emergency") %>% 
  filter(admission_type =="Emergency") %>% 
  group_by(iswinter, year) %>% 
  summarise(count_bad_days = sum(above_thresh), 
            count_days = n()*7, 
            pcnt_bad_days = 100*(sum(above_thresh)/(n()*7)), 
            mean_percentvar =mean(percent_variation)) %>% 
  ggplot() +
  aes(x=year, y=pcnt_bad_days) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(~iswinter)
```


#Section2 - repeat for demographic data

bad days by age group

```{r}
weekly_admissions_demog %>% 
  filter(sex=="All") %>% 
  filter(admission_type =="All") %>% 
  filter(hb_name=="All Scotland") %>% 
  #filter(year == "2021") %>% 
  #filter(iswinter) %>% 
  group_by(age_group) %>% 
  summarise(count_bad_days = sum(above_thresh), 
            count_days = n()*7, 
            pcnt_bad_days = 100*(sum(above_thresh)/(n()*7)), 
            mean_percentvar = mean(percent_variation), 
            #trying a diff calc just to show its the same
            percent_variation = 100*(sum(number_admissions)-sum(average20182019))
                             /sum(average20182019)) 
```
```{r}
weekly_admissions_spec %>% 
  filter(speciality !="All") %>% 
  filter(admission_type =="All") %>% 
  filter(hb_name == "All Scotland") %>% 
  #filter(year =="2022") %>% 
  group_by(speciality) %>% 
  summarise(count_bad_days = sum(above_thresh), 
            count_days = n()*7, 
            pcnt_bad_days = 100*(sum(above_thresh)/(n()*7)), 
            mean_percentvar =mean(percent_variation)) %>% 
  ggplot() +
  aes(x=speciality, y=pcnt_bad_days) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```



```{r}
weekly_admissions_demog %>% 
  filter(sex=="All") %>% 
  filter(admission_type =="All") %>% 
  filter(hb_name=="All Scotland") %>% 
  group_by(age_group) %>% 
  summarise(count_bad_days = sum(above_thresh), 
            count_days = n()*7, 
            pcnt_bad_days = 100*(sum(above_thresh)/(n()*7)), 
            mean_percentvar =mean(percent_variation)) %>% 
  ggplot() +
  aes(x=age_group, y=pcnt_bad_days) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

```{r}
weekly_admissions_demog %>% 
  filter(age_group=="All ages") %>% 
  filter(admission_type =="All") %>% 
  filter(hb_name=="All Scotland") %>% 
  group_by(sex) %>% 
  summarise(count_bad_days = sum(above_thresh), 
            count_days = n()*7, 
            pcnt_bad_days = 100*(sum(above_thresh)/(n()*7)), 
            mean_percentvar =mean(percent_variation)) %>% 
  ggplot() +
  aes(x=sex, y=pcnt_bad_days) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
no obvious difference between sexes

```{r}
weekly_admissions_dep %>% 
  filter(admission_type =="All") %>% 
  filter(hb_name=="All Scotland") %>% 
  group_by(simd_quintile, year) %>% 
  summarise(count_bad_days = sum(above_thresh), 
            count_days = n()*7, 
            pcnt_bad_days = 100*(sum(above_thresh)/(n()*7)), 
            mean_percentvar =mean(percent_variation)) %>% 
  ggplot() +
  aes(x=simd_quintile, y=pcnt_bad_days) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  facet_grid(~year)
```