---
title: "Investigating Data - Covid Wider Impacts"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

Looking at wider covid impacts data
weekly variability - related to 2018/19 baseline

revised version - now extracting all Scotland more easily.

```{r}
library("tidyverse")
library("janitor")
library("lubridate")
```

```{r}
hb_codes <- read_csv("health_board_codes.csv")
hb_codes <- clean_names(hb_codes)
```


```{r}
weekly_admissions <- read_csv("Covid admissions by health board and speciality.csv")
weekly_admissions <- clean_names(weekly_admissions)
```
```{r}
weekly_admissions_demog <- read_csv("Covid admissions by health board, age and sex.csv")
weekly_admissions_demog <- clean_names(weekly_admissions_demog)
```

data cleaning
```{r}
weekly_admissions <- weekly_admissions %>% 
  rename("speciality"= "specialty") %>% 
  rename("speciality_qf"= "specialty_qf") 
```

merge hbnames into data
```{r}
weekly_admissions <- left_join(weekly_admissions,hb_codes) 
```



```{r}
weekly_admissions <- weekly_admissions %>% 
  mutate(year = as.integer(str_sub(week_ending,1,4))) %>% 
  mutate(month = as.integer(str_sub(week_ending,5,6))) %>% 
  mutate(day = as.integer(str_sub(week_ending,7,8))) %>% 
  mutate(wdate = ymd(week_ending)) %>% 
  # identify "All Scotland" data
  mutate(hb_name = ifelse(hb=="S92000003","All Scotland",hb_name)) %>% 
  mutate(hb_name = ifelse(is.na(hb_name),"NHS Region Unknown",hb_name)) %>% 
  mutate(iswinter = ifelse(month %in% c(4,5,6,7,8,9),FALSE,TRUE)) %>% 
  mutate(above_thresh = ifelse(percent_variation>0,7,0))
```

```{r}
weekly_admissions %>% 
  distinct(hb_name)
```

```{r}
weekly_admissions %>% 
  distinct(admission_type)
```

```{r}
weekly_admissions %>% 
  distinct(speciality)
```
calculate totals (just a sense check to see if adding all hb's is the same as "all scotland")
```{r}
total_admissions <- weekly_admissions %>% 
  filter(admission_type == "Emergency") %>% 
  filter(speciality=="All") %>% 
  filter(hb_name != "All Scotland") %>% 
  group_by(wdate) %>% 
  summarise(all_admissions = sum(number_admissions), 
           all_avg20182019 = sum(average20182019)
           ) %>%
  mutate(all_percent_variation = (100*(all_admissions-all_avg20182019)/all_avg20182019)) %>% 
  arrange(desc(all_admissions))
```

Now (just to check - extract same parameter using 'all scotland')
```{r}
total_admissions2 <- weekly_admissions %>% 
  filter(admission_type == "All") %>% 
  filter(speciality=="All") %>% 
  filter(hb_name == "All Scotland") 
```

plot calculated total admissions
```{r}
total_admissions %>% 
ggplot() +
aes(x=wdate, y = all_admissions) +
geom_line(colour='red') +
geom_line(aes(x=total_admissions2$wdate, y = total_admissions2$number_admissions),colour='green', linetype="dotted") 
```
weekly hospital admissions hover around 14000 per week
expect 168,000 per quarter?

```{r}
total_admissions %>% 
ggplot() +  
aes(x=wdate, y = all_avg20182019) +
geom_line(colour='blue') +  
geom_line(aes(x=total_admissions2$wdate, y = total_admissions2$average20182019),colour='green', linetype="dotted") 
```
pre=pandemic hospital admissions hovered around 15500 per week
expect 186,000 per quarter

```{r}
total_admissions %>% 
ggplot() +
aes(x=wdate, y = all_percent_variation) +
geom_line() +
geom_line(aes(x=total_admissions2$wdate, y = total_admissions2$percent_variation),colour='green', linetype="dotted") 
```
Overall(all specialities, all admissions, all healthboards) admissions are only at 90% of pre-pandemic levels

The two lines sit over each other so shows that my calculations were correct.
Can use these calculations when we dont have any averages.


Now look to see which health boards are struggling the most

```{r}
all_admissions_byhb <- weekly_admissions %>% 
  filter(admission_type == "All") %>% 
  filter(speciality=="All") %>% 
  group_by(hb_name) %>% 
  summarise(mean_percnt_var = mean(percent_variation),
            min_percnt_var = min(percent_variation),
            max_percnt_var = max(percent_variation)
           ) %>%
  #mutate(all_percent_variation = all_admissions/all_avg20182019)  
  arrange(desc(max_percnt_var))
all_admissions_byhb
```
```{r}
weekly_admissions %>% 
  filter(admission_type == "All") %>% 
  filter(speciality=="All") %>% 
  group_by(hb_name) %>% 
  ggplot()+
  aes(x=hb_name, y=percent_variation)+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



```{r}
weekly_admissions %>% 
  filter(admission_type == "All") %>% 
  filter(speciality=="All") %>% 
  ggplot() +
  aes(x=wdate, y=percent_variation, group=hb) +
  geom_line()+
  facet_wrap(~hb)
```
What is S92000003 (found answer - its the code for "All Scotland") 
How is NA calculated - still need to check??

```{r}
weekly_admissions %>% 
  filter(admission_type == "All") %>% 
  filter(speciality=="All") %>% 
ggplot() +
aes(x=wdate, y = number_admissions, group = hb) +
geom_line()
```
```{r}
weekly_admissions %>% 
  filter(admission_type == "All") %>% 
  filter(speciality=="All") %>% 
ggplot() +
aes(x=wdate, y = number_admissions, fill=hb) +
geom_col()
```
```{r}
weekly_admissions %>% 
  filter(admission_type == "All") %>% 
  filter(speciality=="All") %>% 
ggplot() +
aes(x=wdate, y = percent_variation, fill=hb) +
geom_col()
```
? what about different departments?
```{r}
weekly_admissions %>% 
  filter(admission_type == "Emergency") %>% 
  filter(speciality=="All") %>% 
  group_by(hb_name) %>% 
  ggplot()+
  aes(x=hb_name, y=percent_variation)+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
weekly_admissions %>% 
  filter(admission_type == "Emergency") %>% 
  filter(speciality=="Accident & Emergency") %>% 
 group_by(hb_name) %>% 
  ggplot()+
  aes(x=hb_name, y=percent_variation)+
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

crisis calculator

```{r}
weekly_admissions %>% 
  filter(admission_type == "All") %>% 
  filter(speciality=="All") %>% 
  #filter(admission_type =="Emergency") %>% 
  #filter(speciality=="Accident & Emergency") %>% 
  #filter(year == 2020) %>% 
  filter(iswinter) %>%
  group_by(hb_name) %>% 
  summarise(pcnt_bad_days = sum(above_thresh)/n()*7, mean_percentvar =mean(percent_variation)) %>% 
  ggplot() +
  aes(x=hb_name, y=mean_percentvar) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#Section2 - repeat for demographic data

