---
title: "Investigating Data - Delayed Discharge"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

Looking at weekly delayed discharge data
what paraeters are available?
https://www.isdscotland.org/health-topics/health-and-social-community-care/delayed-discharges/guidelines/docs/Background-of-delayed-discharge-information-and-glossary-of-terms.pdf

by age 
categories are 18 plus (all ages), 18-74 and 74 +

```{r}
library("tidyverse")
library("janitor")
library("lubridate")
library("snakecase")
library("tsibble")
library("tsibbledata")
```

```{r}
hb_codes <- read_csv("health_board_codes.csv")
hb_codes <- clean_names(hb_codes)
```

setup plots
```{r}
palette = read_csv(here::here("palette_theme/phs_palette.csv"))
# load theme
source(here::here("palette_theme/define_theme.R"))
#load plotting function
source(here::here("palette_theme/plot_timeseries.R"))
```

load in datafiles
```{r}
delayed_discharge <- read_csv("Delayed discharge bed days by health board.csv")
delayed_discharge <- clean_names(delayed_discharge)
```

data cleaning
```{r}
delayed_discharge <- delayed_discharge %>% 
  rename(hb = hbt) %>% #use standardised names
  rename(hbqf = hbtqf) 
```

merge hbnames into datafiles
```{r}
delayed_discharge <- left_join(delayed_discharge,hb_codes)
```

```{r}
delayed_discharge <-delayed_discharge %>% 
  mutate(year = as.integer(str_sub(month_of_delay,1,4))) %>% 
  mutate(month = as.integer(str_sub(month_of_delay,5,6))) %>% 
  mutate(mdate = ym(month_of_delay)) %>% 
  mutate(hb_name = ifelse(hb=="S92000003","All Scotland",hb_name)) %>% 
  mutate(hb_name = ifelse(is.na(hb_name),"NHS Region Unknown",hb_name)) %>%  
  mutate(age_group = ifelse(age_group=="18plus","All (18plus)",age_group)) %>%  
  mutate(iswinter = ifelse(month %in% c(4,5,6,7,8,9),FALSE,TRUE)) 
```

#calculate 2018-2019 levels
```{r}
pre_pandemic_avg <- delayed_discharge %>% 
  filter(between(year,2018,2019)) %>% 
  group_by(hb,age_group,reason_for_delay) %>% 
  summarise(avg_20182019 =  mean(number_of_delayed_bed_days)) 
pre_pandemic_avg
```
#merge this column back into all data
```{r}
delayed_discharge <- left_join(delayed_discharge,pre_pandemic_avg) 
```

```{r}
delayed_discharge <- delayed_discharge %>% 
  mutate(percent_var = 100*(number_of_delayed_bed_days-avg_20182019)/avg_20182019)
```

setup data as a tsibble

```{r}
delayed_discharge <- as_tsibble(delayed_discharge, key = id, index = mdate) 
```

take a look at what is in dataset
```{r}
delayed_discharge %>% 
  distinct(hb_name) 
```
14 health boards and All Scotland

```{r}
delayed_discharge %>% 
  distinct(age_group) 
```
```{r}
delayed_discharge %>% 
  distinct(reason_for_delay) 
```
health and social care reasons - somoene in hospital when they should be in a care home (for example)
patient/family - due to legal disagreements
Code9 - complex reasons
AWI means adults with incapacity


In this dataset we have two parameters we might want to look at.

number_of_delayed_bed_days and average_daily_number_of_delayed_beds
chosen to look at number of delayed beds

look at all data
```{r}
delayed_discharge %>% 
  filter(reason_for_delay == "All Delay Reasons") %>% 
  filter(hb_name == "All Scotland") %>% 
  filter(age_group == "All (18plus)") %>% 
ggplot()+
  aes(x=mdate, y=number_of_delayed_bed_days) +
  geom_line()
```
use function to make this as a standard plot
```{r}
plotdata <- delayed_discharge %>% 
  filter(reason_for_delay == "All Delay Reasons") %>% 
  filter(hb_name == "All Scotland") %>% 
  filter(age_group == "All (18plus)")
plotmapping <- aes(x=mdate, y=percent_var) 
plottitle <- ("Number of delayed bed days - by age group")
plotylabel <- ("% change relative to 2018/19") 
timeseriesplot(plotdata,plotmapping,plottitle,plotylabel)
```

look at data by age group
```{r}
delayed_discharge %>% 
  filter(hb_name =="All Scotland") %>% 
  filter(reason_for_delay == "All Delay Reasons") %>% 
  filter(age_group != "All (18plus)") %>% 
  ggplot() +
  aes(x=mdate, y=number_of_delayed_bed_days, colour=age_group) +
  geom_line()
```
```{r}
delayed_discharge %>% 
  filter(hb_name =="All Scotland") %>% 
  filter(reason_for_delay == "All Delay Reasons") %>% 
  filter(age_group != "All (18plus)") %>% 
  ggplot() +
  aes(x=mdate, y=percent_var, colour=age_group) +
  geom_line()
```

use function to make this as a standard plot
```{r}
plotdata <- delayed_discharge %>% 
  filter(hb_name =="All Scotland") %>% 
  filter(reason_for_delay == "All Delay Reasons") %>% 
  filter(age_group != "All (18plus)")
plotmapping <- aes(x=mdate, y=percent_var, colour=age_group) 
plottitle <- ("Number of delayed bed days - by age group")
plotylabel <- ("% change relative to 2018/19") 
timeseriesplot(plotdata,plotmapping,plottitle,plotylabel)
```


look at data by delay_reason
```{r}
delayed_discharge %>% 
  filter(hb_name =="All Scotland") %>% 
  filter(age_group == "All (18plus)") %>% 
  filter(reason_for_delay != "All Delay Reasons") %>% 
  ggplot() +
  aes(x=mdate, y=number_of_delayed_bed_days, colour=reason_for_delay) +
  geom_line()
```
```{r}
delayed_discharge %>% 
  filter(hb_name =="All Scotland") %>% 
  filter(age_group == "All (18plus)") %>% 
  filter(reason_for_delay != "All Delay Reasons") %>% 
  ggplot() +
  aes(x=mdate, y=percent_var, colour=reason_for_delay) +
  geom_line()
```
use function to make this as a standard plot
```{r}
plotdata <- delayed_discharge %>% 
  filter(hb_name =="All Scotland") %>% 
  filter(age_group == "All (18plus)") %>% 
  filter(reason_for_delay != "All Delay Reasons") 
plotmapping <- aes(x=mdate, y=percent_var, colour=reason_for_delay) 
plottitle <- ("Number of delayed bed days - by reason")
plotylabel <- ("% change relative to 2018/19") 
timeseriesplot(plotdata,plotmapping,plottitle,plotylabel)
```
look at data by delay_reason
```{r}
delayed_discharge %>% 
  filter(reason_for_delay != "All Delay Reasons") %>% 
  filter(hb_name == "All Scotland") %>% 
  filter(age_group == "All (18plus)") %>% 
  filter(year<2020) %>% 
  ggplot() +
  aes(x=iswinter,y=number_of_delayed_bed_days)+
  geom_boxplot() +
  facet_wrap(~year(mdate))
```



```{r}
all_delayed_discharge %>% 
  filter(year>=2020) %>% 
ggplot()+
  aes(x=mdate, y=percent_var) +
  geom_line()
```
```{r}
plotdata <- delayed_discharge %>% 
  filter(reason_for_delay = "All Delay Reasons") %>% 
  filter(hb_name == "All Scotland") %>% 
  filter(age_group == "All (18plus)") #%>% 
  #filter(year>=2020) 
plotmapping <- aes(x=mdate, y=percent_var) 
plottitle <- ("Number of delayed bed days - All Reasons")
plotylabel <- ("% change relative to 2018/19") 
timeseriesplot(plotdata,plotmapping,plottitle,plotylabel)
```

```{r}
plotdata <- delayed_discharge %>% 
  filter(reason_for_delay == "All Delay Reasons") %>% 
  filter(hb_name != "All Scotland") %>% 
  filter(age_group == "All (18plus)") 
plotmapping <- aes(x=mdate, y=percent_var, colour = hb_name) 
plottitle <- ("Number of delayed bed days - by Health Board")
plotylabel <- ("% change relative to 2018/19") 
timeseriesplot(plotdata,plotmapping,plottitle,plotylabel)
```



# investigate variability across health boards


```{r}
discharge_delays_by_hb <- delayed_discharge %>% 
  filter(reason_for_delay == "All Delay Reasons") %>% 
  filter(hb_name != "All Scotland") %>% 
  group_by(mdate, hb_name) %>%
  # summing across all age groups
  summarise(total =  sum(number_of_delayed_bed_days)) %>% 
  mutate(year = year(mdate)) %>% 
  mutate(month = month(mdate)) %>% 
  mutate(iswinter=ifelse(month %in% c(4,5,6,7,8,9),FALSE,TRUE)) 
```
Merge Orkney/Sheltand/WesternIsles
Very low values in these two healthboards make stats analysis difficult
therefore merging them with western isles is recommended.

```{r}
discharge_delays_by_hb %>% 
  group_by(hb_name) %>% 
  summarise(mean = mean(total)) %>% 
  arrange(desc(mean))
```

merge orkney,shetland and western isles
```{r}
discharge_delays_osw <- delayed_discharge %>% 
   pivot_wider(values_from = "total", names_from = "hb_name") 

discharge_delays_by_hb <- clean_names(discharge_delays_by_hb)

discharge_delays_by_hb <- discharge_delays_by_hb %>% 
   mutate(nhs_ork_shet_wi = 
            rowSums(across(c(nhs_orkney,
                             nhs_shetland,
                             nhs_western_isles), na.rm = TRUE))) %>%
   select(-nhs_orkney) %>% 
   select(-nhs_shetland) %>% 
   select(-nhs_western_isles) 
  
discharge_delays_by_hb <- discharge_delays_by_hb %>% 
   pivot_longer(cols = starts_with("nhs"), names_to = "hb_name", values_to = "total") %>%    mutate(hb_name = to_title_case(hb_name)) %>% 
   mutate(hb_name = str_replace(hb_name,"Nhs","NHS"))
```


```{r}
discharge_delays_by_hb  %>% 
  ggplot() +
  aes(x=mdate,y=total,colour = hb_name) +
  geom_line()
```
#calculate 2018-2019 levels
```{r}
pre_pandemic_avg_by_hb <- discharge_delays_by_hb  %>% 
  filter(between(year,2016,2019)) %>% 
  group_by(hb_name) %>% 
  summarise(avg_20182019 =  mean(total, na.rm=TRUE)) 
```
#merge this column back into all data
```{r}
discharge_delays_by_hb <- left_join(discharge_delays_by_hb,pre_pandemic_avg_by_hb) 

discharge_delays_by_hb <- discharge_delays_by_hb %>% 
  mutate(percent_var = 100*((total-avg_20182019)/avg_20182019)) %>% 
  mutate(percent_var = ifelse(total==0,NA_integer_,percent_var))
```

# look at pre-pandemic
was there a summer/winter pattern at hb level?
looking at delayed discharge = higher values is less favourable

```{r}
discharge_delays_by_hb  %>% 
  filter(year<2020) %>% 
  ggplot() +
  aes(x=iswinter,y=percent_var)+
  geom_boxplot() +
  facet_wrap(~hb_name)
```

```{r}
discharge_delays_by_hb  %>% 
  filter(year>=2020) %>% 
  ggplot()+
  aes(x=mdate, y=percent_var, colour = hb_name) +
  geom_line()
```



