---
title: "setting up a plot"
output: html_notebook
---

example code to make a timeseries plot look pretty


```{r}
library("tidyverse")
library("janitor")
library("lubridate")
```

load in datafiles
```{r}
hb_codes <- read_csv("health_board_codes.csv")
hb_codes <- clean_names(hb_codes)
```

```{r}
weekly_admissions_spec <- read_csv("Covid admissions by health board and speciality.csv")
weekly_admissions_spec <- clean_names(weekly_admissions_spec)
```

data cleaning
note probably more cleaning needed before we finalise
I changed the spelling but it turns out both spellings are correct!

```{r}
weekly_admissions_spec <- weekly_admissions_spec %>% 
  rename("speciality"= "specialty") %>% 
  rename("speciality_qf"= "specialty_qf") 
```

merge hbnames into datafiles
```{r}
weekly_admissions_spec <- left_join(weekly_admissions_spec,hb_codes)
```

do the analysis needed for identifying winter and 'crisis'
```{r}
weekly_admissions_spec <- weekly_admissions_spec %>% 
  mutate(year = as.integer(str_sub(week_ending,1,4))) %>% 
  mutate(month = as.integer(str_sub(week_ending,5,6))) %>% 
  mutate(day = as.integer(str_sub(week_ending,7,8))) %>% 
  mutate(wdate = ymd(week_ending)) %>% 
  # identify "All Scotland" data
  mutate(hb_name = ifelse(hb=="S92000003","All Scotland",hb_name)) %>% 
  mutate(hb_name = ifelse(is.na(hb_name),"NHS Region Unknown",hb_name)) %>% 
  mutate(iswinter = ifelse(month %in% c(4,5,6,7,8,9),FALSE,TRUE)) %>% 
  mutate(above_thresh = ifelse(percent_variation>0,7,0)) %>% 
  filter(between(year,2020,2022)) # we are not interested in any 2023 data
```


```{r}
weekly_admissions_spec %>% 
  filter(admission_type == "All") %>% 
  filter(speciality == "All") %>% 
  filter(hb_name != "All Scotland") %>% 
ggplot() +
aes(x=wdate, y = number_admissions, colour = hb_name) +
geom_line()
```


```{r}
palette = read_csv(here::here("palette_theme/phs_palette.csv"))
# load theme
source(here::here("palette_theme/define_theme.R"))
#load plotting function
source(here::here("palette_theme/plot_timeseries.R"))
```
#example of creating the basic plot
```{r}
weekly_admissions_spec %>% 
  #filter(admission_type == "All") %>% 
  filter(speciality == "All") %>% 
  filter(hb_name == "All Scotland") %>% 
ggplot() +
aes(x=wdate, y = number_admissions, colour = admission_type) +
geom_line() +
scale_colour_manual(values = palette$mycolours) 
```
#example of how it would look when you use the function

```{r}
plotdata <- weekly_admissions_spec %>% 
  #filter(admission_type == "All") %>% 
  filter(speciality == "All") %>% 
  filter(hb_name == "All Scotland") 
plotmapping <- aes(x=wdate, y = number_admissions, colour = admission_type)
plottitle <- ("Hospital Admissions of people with Covid")
plotylabel <- ("Number of Hospital Admissions") 
timeseriesplot(plotdata,plotmapping,plottitle,plotylabel) +
  # as long as you include a + in the line above you can make further modifications
  #ggtitle("change the title if you want to")+
  #e.g add extra lines if you want to
  geom_hline(yintercept=12000, linewidth = 0.5) 
```





