---
title: "setting up a plot"
output: html_notebook
---

example code to make a timeseries plot look pretty


```{r}
library("tidyverse")
library("janitor")
library("lubridate")
```

load in datafiles
```{r}
hb_codes <- read_csv("health_board_codes.csv")
hb_codes <- clean_names(hb_codes)
```

```{r}
weekly_admissions_spec <- read_csv("Covid admissions by health board and speciality.csv")
weekly_admissions_spec <- clean_names(weekly_admissions_spec)
```

data cleaning
note probably more cleaning needed before we finalise
I changed the spelling but it turns out both spellings are correct!

```{r}
weekly_admissions_spec <- weekly_admissions_spec %>% 
  rename("speciality"= "specialty") %>% 
  rename("speciality_qf"= "specialty_qf") 
```

merge hbnames into datafiles
```{r}
weekly_admissions_spec <- left_join(weekly_admissions_spec,hb_codes)
```

do the analysis needed for identifying winter and 'crisis'
```{r}
weekly_admissions_spec <- weekly_admissions_spec %>% 
  mutate(year = as.integer(str_sub(week_ending,1,4))) %>% 
  mutate(month = as.integer(str_sub(week_ending,5,6))) %>% 
  mutate(day = as.integer(str_sub(week_ending,7,8))) %>% 
  mutate(wdate = ymd(week_ending)) %>% 
  # identify "All Scotland" data
  mutate(hb_name = ifelse(hb=="S92000003","All Scotland",hb_name)) %>% 
  mutate(hb_name = ifelse(is.na(hb_name),"NHS Region Unknown",hb_name)) %>% 
  mutate(iswinter = ifelse(month %in% c(4,5,6,7,8,9),FALSE,TRUE)) %>% 
  mutate(above_thresh = ifelse(percent_variation>0,7,0)) %>% 
  filter(between(year,2020,2022)) # we are not interested in any 2023 data
```


```{r}
weekly_admissions_spec %>% 
  filter(speciality == "All") %>% 
  filter(admission_type == "All") %>% 
  filter(hb_name =="All Scotland") %>% 
ggplot() +
aes(x=wdate, y = number_admissions) +
geom_line() 
```
setup theme for plots
```{r}
palette = read_csv("phs_palette.csv")

theme_phs <- 
    theme(
        text = element_text(size = 14),
        title = element_text(size = 14),
        axis.text = element_text(size = 14),
        legend.text = element_text(size = 12),
        # work out how to make graph transparent?
        panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill = "transparent", colour = NA),
        legend.box.background = element_rect(fill = "transparent", colour = NA),
        panel.grid = element_line(colour = "grey90", linetype = "dashed"),
        legend.position = "bottom"
    )
```

```{r}
plotlim <- as.Date(c("2020-01-01","2022-12-31")) 
p <-  ggplot(
  weekly_admissions_spec %>% 
  filter(speciality == "All") %>% 
  filter(admission_type == "All") %>% 
  filter(hb_name =="All Scotland")  
  ) +
aes(x=wdate, y = number_admissions) +
geom_line(color = palette$colours[2]) +    
#add dashed lines to show the boundaries 
source(here::here(""))
scale_x_date(limits=plotlim, date_breaks="3 month", labels = scales::label_date_short(), expand = c(0,0.01)) +
#scale_y_continuous(expand = c(0,0)) +
xlab("Date") +   
ylab("Hospital Admissions with Covid") +
theme_phs 
p
#extract ylimits from plot
ylimits = layer_scales(p)$y$get_limits()
xlimits = layer_scales(p)$x$get_limits()
#add gray shaded rectangles to note winter
p + annotate("rect", 
          xmin=c(ymd(plotlim[1]),ymd(20201001),ymd(20211001),ymd(20221001)),
          xmax=c(ymd(20200401),ymd(20210401),ymd(20220401),plotlim[2]), 
         ymin=c(-Inf,-Inf,-Inf,-Inf), 
         ymax=c(Inf,Inf,Inf,Inf), 
         alpha=0.1, fill=palette$colours[4]) 
```
## How would we map a plot?

add_winterlines <- function(p, strings){
 p + geom_vline(xintercept=ymd(20201001),color="gray",linetype="dotted", size = 1) +   
 p + geom_vline(xintercept=ymd(20211001),color="gray",linetype="dotted", size = 1) + 
 p +  geom_vline(xintercept=ymd(20221001),color="gray",linetype="dotted", size = 1 ) + 
geom_vline(xintercept=ymd(20200401),color="gray",linetype="dotted", size = 1) + 
geom_vline(xintercept=ymd(20210401),color="gray",linetype="dotted", size = 1) + 
geom_vline(xintercept=ymd(20220401),color="gray",linetype="dotted", size = 1) + 



```{r}
library(sf)
map_scot <- st_read("SG_NHS_HealthBoards_2019/SG_NHS_HealthBoards_2019.shp")
tx_ll <- st_transform(map_scot, "+proj=longlat +ellps=WGS84 +datum=WGS84")

ggplot(map_scot) +
 geom_sf() +
 stat_sf_coordinates(geom = "point", color = "red") +
 geom_sf_text(aes(label = HBName), size = 3, nudge_x = 600, colour = "black",check_overlap = TRUE) +
 scale_fill_distiller(palette = "GnBu", type = "seq") +
 theme_void()
```
reminder for myself of of how I plotted long lat onto map before...

st_stations <- data.frame(lat = c(40.73,40.73,40.709,40.709,40.73),
                               long = c(-74.052,-74.03,-74.03,-74.052,-74.052))
stations_stf <- st_stations %>% 
  st_as_sf(coords = c("long","lat"), crs = 4326) 


extracting colours from palette
```{r}
library(colorfindr)
## setting up color palette
# setup colour scheme
# get colours from citibike logo
phspal <- get_colors("PHSlogo.png") %>%  
make_palette(n = 6)
colours <- phspal # If you rerun this colour selection will change
phspal <- as.data.frame(colours)

#write_csv(phspal,"phs_palette.csv")
```


