---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(skimr)
```


Example of creating summary indicators by health board

# note you dont need this read script as its already read into the app.
```{r}
admissions_spec <- read_csv(here::here("clean_data/weekly_admissions_spec_clean.csv")) 
delayed_discharge <- read_csv(here::here("clean_data/delayed_discharge_clean.csv"))  
```


# extract admissions data (Emergency)
```{r}
# extract 2018/19 average and current value
seldata_admissions <- admissions_spec %>% 
  filter(admission_type == "Emergency") %>% 
  filter(speciality=="All") %>% 
  filter(wdate == max(wdate)) %>% 
  #group_by(hb_name) %>% 
  select(hb_name, average20182019, number_admissions, percent_var) 
  
# add this to summary table
ha1_byhb <- seldata_admissions %>% 
  filter(hb_name != "All Scotland") 

# add this to plot
ha1_tot <- seldata_admissions %>% 
 filter(hb_name =="All Scotland")
```


# extract admissions data (Planned)
```{r}
# extract 2018/19 average and current value
seldata_admissions <- admissions_spec %>% 
  filter(admission_type == "Planned") %>% 
  filter(speciality=="All") %>% 
  filter(wdate == max(wdate)) %>% 
  #group_by(hb_name) %>% 
  select(hb_name, average20182019, number_admissions, percent_var) 
  
# add this to summary table
ha2_byhb <- seldata_admissions %>% 
  filter(hb_name != "All Scotland") 

# add this to plot
ha2_tot <- seldata_admissions %>% 
 filter(hb_name =="All Scotland")
```

#extract delayed discharge data
# extract 2018/19 average and current value
```{r}
seldata <- delayed_discharge %>% 
  filter(age_group == "All (18plus)") %>% 
  filter(reason_for_delay=="All Delay Reasons") %>% 
  filter(mdate == max(mdate)) %>% 
  #group_by(hb_name) %>% 
  select(hb_name, avg_20182019, number_of_delayed_bed_days, percent_var) 
  
# add this to summary table
dd_byhb <- seldata %>% 
  filter(hb_name != "All Scotland") 

# add this to plot
dd_tot <- seldata %>% 
 filter(hb_name =="All Scotland")
```


# parameter names
```{r}
"Admissions to Hospital - Emergency"
"Admissions to Hospital - Planned"
"Delayed Discharge Bed Days"
```


# values you need for map are
```{r}
ha1_byhb
ha2_byhb
dd_byhb
```
#values you need for table are
```{r}
ha1_tot
ha2_tot
dd_tot
```

Converting bed days to percentage occupied by delayed patients

```{r}
num_beds <- read_csv(here::here("clean_data/num_beds_by_hb.csv"))

# multiple by 30 so we have number of beds days available by month for each   
# health board

num_beds %>% 
  mutate(num_bed_days_each_month = num_beds * 30) %>% 
  cbind(dd_byhb$number_of_delayed_bed_days) %>% 
  rename(number_of_delayed_bed_days = "dd_byhb$number_of_delayed_bed_days") %>% 
  mutate(percent_bed_days_taken_by_delayed_discharge = 
           number_of_delayed_bed_days / num_bed_days_each_month * 100)


```
```{r}




```

